---
title: "TBL (Week 4) Questions"
author: "Ruoqing Zhu"
date: "Last Updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: hide
    df_print: paged
    toc: yes
    toc_float:
      collapsed: true
      smooth_scroll: true
    toc_depth: '2'
  pdf_document:
    toc: yes
    toc_depth: 2
---

```{r set-options, echo=FALSE, cache=FALSE}
  options(width = 1000)
  knitr::opts_chunk$set(fig.width=9, fig.height=6, out.width = "75%", fig.align = 'center')
  knitr::opts_chunk$set(class.source = "fold-show")
  knitr::opts_chunk$set(collapse=FALSE)
```
```{css, echo=FALSE}
.questioncode {
background-color: #CCDDFF;
}
```

## Questions 1

In 1852, William Farr published a report of the Registrar-General on mortality due to cholera in England in the years 1848-1849. Farr believed that cholera arose from bad air "miasma" associated with low elevation above the River Thames. The data collected by Farr is provided in the `HistData` package. Use the following code to load the data 

```{r}
  library(HistData)
  data(Cholera)
  write.table(data.frame(Cholera), "Cholera.txt")
```

In this data, `cholera_drate` is the death rate, and we will perform correlation analysis with the variable `elevation`. What is the Pearson correlation between these two variables?

```{r}
  cor(Cholera$cholera_drate, Cholera$elevation)
  cor(Cholera$cholera_deaths, Cholera$elevation)
  cor(Cholera$popn, Cholera$elevation)
  cor(Cholera$annual_deaths, Cholera$elevation)
```

  * `A` -0.4823152
  * `B` -0.3819684
  * `C` -0.0425442
  * `D` -0.3068787
  
`A`

## Questions 2 
  
Test the significance of this correlation. What is the p-value? 

```{r}
  cor.test(Cholera$cholera_drate, Cholera$elevation)$p.value
  cor.test(Cholera$cholera_deaths, Cholera$elevation)$p.value
  cor.test(Cholera$popn, Cholera$elevation)$p.value
  cor.test(Cholera$annual_deaths, Cholera$elevation)$p.value
```

  * `A` 0.0021655
  * `B` 0.0179585
  * `C` 0.7997937
  * `D` 0.0609254

`A`

## Questions 3

The following is the output from a Pearson correlation and the associated test from another dataset. Which statement is correct?

```{r class.source="questioncode"}
  x = rnorm(1000)
  y = 0.1*x + rnorm(1000)
  cor(x, y)
  cor.test(x, y)
```

  * `A` The correlation between $x$ and $y$ is very weak
  * `B` The the correlation between $x$ and $y$ is quite strong since it is highly significant (0.001)
  * `C` The chance that $x$ and $y$ are not correlated is less than 0.001
  * `D` The chance that the true correlation falls in between 0.0398 and 0.1625 is 95%

The correct answer is `A` since significance and hypothesis testings cannot be interpreted as probabilities. 

## Questions 4

Perform a linear regression that uses the death rate as the outcome and elevation as the covariate. Answer two questions: 1) For each unit change of elevation, what is the corresponding change of death rate; 2) Is elevation a significant variable if we use an alpha level at 0.001? 

```{r}
  fit = lm(cholera_drate ~ elevation, data = Cholera)
  summary(fit)
```

  * `A` Increase by 0.4388; Yes
  * `B` Decrease by 0.4388; Yes
  * `C` Increase by 0.4388; No
  * `D` Decrease by 0.4388; No  

`D`. Since 0.00217 is larger than 0.001.

## Questions 5

Base on this regression model, predict the death rate when the elevation is 80.

```{r}
  newdata = data.frame("elevation" = 80)
  predict(fit, newdata)
```

  * `A` 48.70548
  * `B` 83.8117
  * `C` 101.3637
  * `D` 39.9317

`A`

## Questions 6

Add a new covariate `water` into the previous linear regression, and treat it as a nominal categorical variable. Answer two questions: 1) How many categories does water have; 2) Is elevation significant (at 0.05 level)? 

```{r}
  summary(lm(cholera_drate ~ as.factor(water) + elevation, data = Cholera))
```

  * `A` 2; Yes
  * `B` 3; Yes
  * `C` 2; No
  * `D` 3; No 

`D`

## Questions 7

If you noticed the change of significance (at 0.05 level) of elevation. What do you think is the cause? 

  * `A` Different water supply regions are at very different elevation level, making the two variables essentially co-linear.
  * `B` The water supply regions variable is taking effects away from elevation, making it insignificant
  * `C` This is a confounding variable issue and water supply regions is the true cause
  * `D` All of the others 
  
`D`

## Questions 8

A researcher decide to add the poor rate (`poor_rate`, a tax on property levied in each parish) into the model. This is a continuous variable. Perform this linear model using R. Which of the following conclusions is correct?

```{r}
  summary(lm(cholera_drate ~ as.factor(water) + elevation + poor_rate, data = Cholera))  
```
  * `A` The poor rate variable explained a significant amount of variation of the death rate
  * `B` The poor rate variable did not make much improvement since the F-statistic $p$ value remained similar
  * `C` The poor rate variable is more significant than all other variables
  * `D` The effect of poor rate is much larger than other variables because its estimated parameter is much larger
  
`A`

## Questions 9

Read Tosepu et al. (2020), which was provide in the practice question of the reading material. What can be the potential issues that makes the conclusion invalid, from a statistical point of view? 

  * `A` The sample size is too small
  * `B` The model does not control for enough factors that may potentially affect the infection rate 
  * `C` The cases could be underestimated since people may not get tested for COVID
  * `D` All of the others

`D`

