---
title: "TBL (Week 4) Questions"
author: "Ruoqing Zhu"
date: "Last Updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: hide
    df_print: paged
    toc: yes
    toc_float:
      collapsed: true
      smooth_scroll: true
    toc_depth: '2'
  pdf_document:
    toc: yes
    toc_depth: 2
---

```{r set-options, echo=FALSE, cache=FALSE}
  options(width = 1000)
  knitr::opts_chunk$set(fig.width=9, fig.height=6, out.width = "75%", fig.align = 'center')
  knitr::opts_chunk$set(class.source = "fold-show")
  knitr::opts_chunk$set(collapse=FALSE)
```
```{css, echo=FALSE}
.questioncode {
background-color: #CCDDFF;
}
```

## Questions 1

The following is the output from a Pearson correlation and the associated test. Which statement is correct?

```{r class.source="questioncode"}
  x = rnorm(1000)
  y = 0.1*x + rnorm(1000)
  cor(x, y)
  cor.test(x, y)
```

  * `A` The correlation between $x$ and $y$ is very weak
  * `B` The the correlation between $x$ and $y$ is quite strong since it is highly significant (0.001)
  * `C` The chance that $x$ and $y$ are not correlated is less than 0.001
  * `D` The probability that the correlation falls in between 0.0398 and 0.1625 is 95%

The correct answer is `A` since significance and hypothesis testings cannot be interpreted as probabilities. 

## Questions 2

```{r}
  library(HistData)
  data("Cholera")
  summary(lm(cholera_drate ~ as.factor(water) + elevation, data = Cholera))
  
  summary(lm(cholera_drate ~ elevation, data = Cholera))
  cor(Cholera$elevation, Cholera$cholera_drate)
```

