---
title: "Week 3 In-class Practice"
author: "Ruoqing Zhu"
date: "Last Updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: hide
    df_print: paged
    toc: yes
    toc_float:
      collapsed: true
      smooth_scroll: true
    toc_depth: '2'
  pdf_document:
    toc: yes
    toc_depth: 2
---

```{r set-options, echo=FALSE, cache=FALSE}
  options(width = 1000)
  knitr::opts_chunk$set(fig.width=9, fig.height=6, out.width = "75%", fig.align = 'center')
  knitr::opts_chunk$set(class.source = "fold-show")
  knitr::opts_chunk$set(collapse=TRUE)
```

## Correlation

1. Estimate and test the significance of the correlation between the father height and mother height in the `GaltonFamilies` dataset. Do you think the two variables are independent?
  
```{r class.source = NULL, eval = FALSE}
  cor(GaltonFamilies$father, GaltonFamilies$mother)
  cor.test(GaltonFamilies$father, GaltonFamilies$mother)
```


2. Read Table 1 in [Tosepu et al. (2020)](https://teazrq.github.io/Stat-PopHealth/papers/Tosepu2020.pdf). Which variable is considered significantly correlated with the COVID19 cases count? What type of correlation is used in this paper? And why? What could be potential issues that make the conclusion invalid?

## Simple Linear Regression 

  1. Fit a simple linear regression that uses father's height to predict the Child's height using the `GaltonFamilies` dataset. 
      * For each inch increase of father's height, how much inch the child is expected to increase?
      * Is this variable significant?
      * What is the R-squared from this model? And from this information, what is the correlation between the fitted value and the outcome?
      * Predict a new subject with father's height being 70 inches. 
  
```{r class.source = NULL, eval = FALSE}
  summary(lm(childHeight~father, data = GaltonFamilies))
```
  
  2. Based on your understanding of a simple linear regression, is this model a good fitting? 
  
  3. Read Table 2 in [Goodings et al. (2011)](https://teazrq.github.io/Stat-PopHealth/papers/Gooding2011.pdf), what is the p-value associated with food insecure when modeling BMI in females using a simple linear regression? 


## Multiple Linear Regression 


  1. Fit a multiple linear regression models the Child's height using the `GaltonFamilies` dataset. You should consider the following terms in your regression:
  
      * Father's height
      * Mother's height
      * An interaction between father and mother's heights
      * Number of children in the family, as a nominal categorical variable 
    
```{r class.source = NULL, eval = FALSE}
  library(HistData)
  data(GaltonFamilies)
  summary(lm(childHeight ~ father + mother + I(father*mother) + as.factor(children), data = GaltonFamilies))
```

  2. Do you notice any problem with this model? What could be the causes? Now, instead of defining the number of children as a nominal categorical variable, treat it as an ordinal one (continuous). What is the most significant variable in this new model?
  
```{r class.source = NULL, eval = FALSE}  
  summary(lm(childHeight ~ father + mother + I(father*mother) + children, data = GaltonFamilies))
```

  3. The result still contradict with our understanding that both father and mother's heights would affect a child's height. Remove the interaction term and refit the model. What do you think is the cause of this change? Based on this model, predict a child's height if the father is 75, mother is 65 and there are 3 children in this family. 
  
```{r class.source = NULL, eval = FALSE} 
  summary(lm(childHeight ~ father + mother + children, data = GaltonFamilies))
```

  4. Can you think of examples in which not including a confounding variable would cause a misleading result? 

  5. Read Table 2 in [Goodings et al. (2011)](https://teazrq.github.io/Stat-PopHealth/papers/Gooding2011.pdf) and contrasting Model 2 and Model 3. What could be the cause that the category `Multiethnic` parameter estimate changed quite significantly?



