---
title: "TBL Reading (Week 4): Risk Ratio and Odds Ratio"
author: "Ruoqing Zhu"
date: "Last Updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: hide
    df_print: paged
    toc: yes
    toc_float:
      collapsed: true
      smooth_scroll: true
    toc_depth: '2'
  pdf_document:
    toc: yes
    toc_depth: 2
---

```{r set-options, echo=FALSE, cache=FALSE}
  options(width = 1000)
  knitr::opts_chunk$set(fig.width=9, fig.height=6, out.width = "75%", fig.align = 'center')
  knitr::opts_chunk$set(class.source = "fold-show")
  knitr::opts_chunk$set(collapse=TRUE)
```

## Overview 



## Contengency Table 




## Practice questions 

  1. Estimate and test the significance of the correlation between the father height and mother height in the `GaltonFamilies` dataset. Do you think the two variables are independent?
  
```{r class.source = NULL, eval = FALSE}
  cor(GaltonFamilies$father, GaltonFamilies$mother)
  cor.test(GaltonFamilies$father, GaltonFamilies$mother)
```

  2. Read Table 1 in [Tosepu et al. (2020)](papers/Tosepu2020.pdf). Which variable is considered significantly correlated with the COVID19 cases count? What type of correlation is used in this paper? And why?



