---
title: "Week 1 In-class Practice"
author: "Ruoqing Zhu"
date: "Last Updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: hide
    df_print: paged
    toc: yes
    toc_float:
      collapsed: true
      smooth_scroll: true
    toc_depth: '2'
  pdf_document:
    toc: yes
    toc_depth: 2
---

```{r set-options, echo=FALSE, cache=FALSE}
  options(width = 1000)
  knitr::opts_chunk$set(fig.width=9, fig.height=7, out.width = "75%", fig.align = 'center')
  knitr::opts_chunk$set(class.source = "fold-show")
  knitr::opts_chunk$set(collapse=TRUE)
```

## Overview 

Before running this file, you should already have R and Rstudio installed. In addition, the first time you use this file, RStuido would suggest you to install RMarkdown, which is a tool to compile this file to a nice looking pdf or html file. This is useful if you want to document your work for future use. But of course, this is not a must. 

The follow practice questions represent an overly simplified process of data analysis. 

## Load Data

Load the iris data from R so that we can perform further analysis 

```{r class.source = NULL, eval = FALSE}
  data(iris)
```

## Summary Statistics

Calculate the mean of variable `Petal.Length` in the iris data. 

```{r class.source = NULL, eval = FALSE}
  mean(iris$Petal.Length)
```

Calculate the median of variable `Petal.Width` in the iris data. 

```{r class.source = NULL, eval = FALSE}
  median(iris$Petal.Length)
```

## Subsetting Data

Calculate the mean of variable `Petal.Length` for observations of `setosa` species.

```{r class.source = NULL, eval = FALSE}
  mean(iris$Petal.Length[iris$Species == "setosa"])
```

## Data Visualization

Produce a scatter plot of `Petal.Length` vs. `Petal.Width`.

```{r class.source = NULL, eval = FALSE}
  plot(iris$Petal.Length, iris$Petal.Width)
```

## Creating New Variables

Attach a new numerical column to the `iris` data, as the product of `Petal.Length` and `Petal.Width` and name the column as `Petal.Prod`.

```{r class.source = NULL, eval = FALSE}
  iris = cbind(iris, "Petal.Prod" = iris$Petal.Length*iris$Petal.Width)
  head(iris)
```
