---
title: "TBL (Week 5) Questions"
author: "Ruoqing Zhu"
date: "Last Updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: hide
    df_print: paged
    toc: yes
    toc_float:
      collapsed: true
      smooth_scroll: true
    toc_depth: '2'
  pdf_document:
    toc: yes
    toc_depth: 2
---

```{r set-options, echo=FALSE, cache=FALSE}
  options(width = 1000)
  knitr::opts_chunk$set(fig.width=9, fig.height=6, out.width = "75%", fig.align = 'center')
  knitr::opts_chunk$set(class.source = "fold-show")
  knitr::opts_chunk$set(collapse=FALSE)
```
```{css, echo=FALSE}
.questioncode {
background-color: #CCDDFF;
}
```

## Questions 1

Use the following code to load a sepsis dataset from the `aVirtualTwins` package. You may need to install this package prior to loading. This is a dataset with 470 patients and 13 variables. 

```{r}
  library(aVirtualTwins)
  data(sepsis)
```

The outcome is survival after 28 days of treatment, a value of 1 for subjects who died by 28 days and 0
otherwise. The Therapy variable is an indicator of a newly developed treatment (with value 1) versus the control (value 0). The following code can be used to summarize the observed data into a contingency table. 

```{r}
  table("Therapy"= sepsis$THERAPY, "Survival" = sepsis$survival)
```

Assuming that this is a randomized trial, what is the relative risk of death associated with using the new therapy, compared with the control? 

    [A]. 1.197343
    B. 0.835182
    C. 0.898397
    D. 1.113093

```{r}
  library(epitools)
  freqtable = table("Therapy"= sepsis$THERAPY, "Survival" = sepsis$survival)
  riskratio(freqtable)
```
```{r class.source = NULL, eval = FALSE}
  # wrong answers
  riskratio(freqtable[c(2,1), ])$measure
  riskratio(freqtable[, c(2,1)])$measure
  riskratio(freqtable[c(2,1), c(2,1)])$measure
```

## Questions 2

Based on the result from the previous question, the new treatment is 

    A. Significantly better than control 
    B. Significantly worse than control
    [C]. not siginificantly different from control 

## Questions 3

Now, we will use the McLaughlin et al. (2003) Table 2 to perform the analysis. You need to setup the contingency table yourself based on the information provided in the paper and answer the next several questions. Consider the CD4 cell count $\geq$ 200 as the indication of exposure and use infection as the outcome of interest. What is the 95% confidence interval of the odds ratio corresponding to this exposure?

```{r}
  freqtable = matrix(c(11, 118, 15, 89),nrow = 2, ncol = 2)
  oddsratio(freqtable)
```
    [A]. 0.2362064, 1.272439
    B. 0.7858923, 4.233585
  
```{r}
  oddsratio(freqtable[c(2,1), ])$measure
```

## Question 4 

If we switch the explore to CD4 cell count $<$ 200 and also switch the outcome of interest to not infected. Then the odds ratio would be

    A. The reciprocal of the previous one
    [B]. Unchanged
    C. Increase because CD4 cell can trigger immune cells to protect against infection 
    D. None of the others
  
## Question 5 

Following the results from Question 3, high CD4 cell count can 

    A. Significantly reduce the risk of infection  
    B. Significantly increase the risk of infection
    [C]. not significantly  different from low CD4 cell count 

## Question 6

Which of the following study design is suitable for calculating the relative risk, but not suitable for odds ratio?

    A. A randomized clinical trial
    B. Case-control study of cancer-associated genetic marker
    C. Observational study of the accuracy of a COVID19 test kit
    [D]. None of the others
    
## Question 7

Using the sepsis data again, fit a logistic regression to predict the probability of death (survival = 1) using THERAPY as a binary predictor and AGE as a continuous predictor. 

```{r}
  logistic.fit <- glm(survival~THERAPY + AGE, data = sepsis, family = binomial)
  summary(logistic.fit)
```

Regardless of significance or not, if a patient is 50 years old and is using the new therapy, what is the probability of death? 

    A. -0.842635
    B. 0.4305745
    C. 0.0523307
    [D]. 0.3009837

```{r}
  -3.206481 + 0.268146 + 50*0.041914
  testdata = data.frame("THERAPY" = 1, AGE = 50)
  predict(logistic.fit, newdata = testdata, type = "response")
```

## Question 8

Based on the parameter estimate, each unit increase of Age would

    A. Increase the predicted probability of death by 0.041914
    B. Increase log of the predicted probability of death by 0.041914
    C. Increase the odds ratio of death by 0.041914
    [D]. Increase log of the odds ratio of death by 0.041914

## Question 9

Using experiment 1 in Table 1 of O’Brien (2021), there are five parameter estimates based on the reduced model (the right four columns). Column B is the estimated parameter. The model involves an interaction term between Scientific Content and Trust. Consider a new sample, with scientific content = 1. If we increase trust by one unit, then 

    A. The log odds ratio of disseminating the information would decrease by 0.5
    [B]. The log odds ratio of disseminating the information would decrease by 0.02
    C. The probability of disseminating the information would decrease by 0.5
    D. The probability of disseminating the information would decrease by 0.02







