---
title: "Testing Mean Differences and Associations"
author: "Ruoqing Zhu"
date: "Last Updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: hide
    df_print: paged
    toc: yes
    toc_float:
      collapsed: true
      smooth_scroll: true
    toc_depth: '2'
  pdf_document:
    toc: yes
    toc_depth: 2
---

```{r set-options, echo=FALSE, cache=FALSE}
  options(width = 1000)
  knitr::opts_chunk$set(fig.width=9, fig.height=6, out.width = "75%", fig.align = 'center')
  knitr::opts_chunk$set(class.source = "fold-show")
  knitr::opts_chunk$set(collapse=TRUE)
```

## Discussion p-values
  
  * If we get $p$-value $=0$ (in certain situations we could), can we know for sure that the Null hypothesis is wrong?
  * Based on the testing procedure, what is the chance that we make a mistake when the alternative hypothesis is true?
  * Discussion: What factors could affect the type II error?
  
## z-test

Following the same assumptions as the our z-test question, suppose we are interested in testing whether the average score is **smaller than 95**. Can you obtain the $p$-value of this test?

```{r class.source = NULL, eval = FALSE}
  pnorm(90 - 95, sd = 15/sqrt(48), lower.tail = TRUE)
```

## One Sample t-test

If we only get to observe the first 8 observations of the Age variable. Perform the same hypothesis testing problem, what is your conclusion? What is affecting the results? Discuss the impact of having a small sample size.

```{r}
  heart = read.csv("processed_cleveland.csv")
```

```{r class.source = NULL, eval = TRUE}
    t.test(heart$age[1:8], mu = 50, alternative = "greater")
```

