---
title: "Read-in NHANES Data"
author: "Ruoqing Zhu"
date: "Last Updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '2'
  pdf_document:
    toc: yes
    toc_depth: 2
---

```{r set-options, echo=FALSE, cache=FALSE}
  options(width = 1000)
  knitr::opts_chunk$set(fig.width=9, fig.height=7, out.width = "60%", fig.align = 'center')
  knitr::opts_chunk$set(collapse=TRUE)
```

## The NHANES Data

I have already downloaded a bunch of `.XPT` files from the CDC NHANES dataset (see their links below) and saved them under the same folder as this `.rmd` file. The following code is an example for processing them so that its ready to analyze. Please note that you will not be able to replicate this unless you download all the following `.XPT` files from [CDC NHANES website](https://wwwn.cdc.gov/nchs/nhanes/) and save them into the same folder as this `ReadNHANESData.Rmd` file. For demonstration purpose, I will create a toy dataset `NHANCE_2016_Toy.Rdata` to perform some calculations. If you plan to use the NHANCE data in your final project, you should consider reproducing this with additional variables (from more `.XPT` files) that you are interested. 

## Readin Data

For example, I have the following setup in my folder. To reproduce this, you should have all the `.XPT` files, and the `.Rmd` file provided at our [course website](https://teazrq.github.io/Stat-PopHealth/week2.html). R and R Markdown tutorial can be found on the same page. The following code chuck is used to read-in data from `.XPT` files. 

```{r}
  # this is a package used to read .XPT files
  # if you don't have the package installed already, you should run 
  # install.packages("SASxport") first

  library(SASxport)
```

```{r}  
  # Read-In HNANES 2015 - 2016 data 
  # https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?BeginYear=2015

  # Demographics Data
  Demographic <- read.xport("DEMO_I.XPT")         
    # https://wwwn.cdc.gov/Nchs/Nhanes/2015-2016/DEMO_I.XPT
  
  # Dietary Data
  DietaryFood1 <- read.xport("DR1IFF_I.XPT")      
    # https://wwwn.cdc.gov/Nchs/Nhanes/2015-2016/DR1IFF_I.XPT

  # Questionnaire Data
  Diabetes <- read.xport("DIQ_I.XPT")             
    # https://wwwn.cdc.gov/Nchs/Nhanes/2015-2016/DIQ_I.XPT
  
  # Examination Data
  BodyMeasure <- read.xport("BMX_I.XPT")          
    # https://wwwn.cdc.gov/Nchs/Nhanes/2015-2016/BMX_I.XPT

  # You can download more data from the CDC website
```

## Processing the data

```{r}
  # We consider a few variables 

  Demographic_Sub = Demographic[, c("SEQN", "RIDAGEYR")]
  Diabetes_Sub = Diabetes[, c("SEQN", "DIQ010")]
  BodyMeasure_Sub = BodyMeasure[, c("SEQN", "BMXBMI")]
  Food1_Sub = DietaryFood1[, c("SEQN", "DR1ISUGR", "DR1IKCAL")]
  Food1_Sub = aggregate(. ~ SEQN, Food1_Sub, FUN=sum)
  
  # Merge all dataset by SEQN (the unique identifier)
  # This is a rather complicated function
  RawData = Reduce(function(x,y) merge(x = x, y = y, by = "SEQN", all.x = TRUE),
                   list(Demographic_Sub, Diabetes_Sub, Food1_Sub, BodyMeasure_Sub))
  
  RawData = RawData[complete.cases(RawData), ]
```

For convenience, you should save the processed data so that you do not need to go through that again. 

```{r, fig.height=5, fig.width=7, message=FALSE}
  # Save data
  save(RawData, file = "NHANCE_2016_Toy.Rdata")
```

## Using the data for analysis

The following code will load and view the data you just saved

```{r, fig.height=5, fig.width=7, message=FALSE}
  # Save data
  load("NHANCE_2016_Toy.Rdata")
  head(RawData)
```

Now you can perform analysis using this data. For example, 

```{r, fig.height=5, fig.width=7, message=FALSE}
  # Get the mean of age 
  mean(RawData$RIDAGEYR)

  # Plot calorie intake and BMI, colored by diabetes status
  library(ggplot2)
  ggplot(RawData, aes(DR1IKCAL, BMXBMI, colour = as.factor(DIQ010))) + 
    geom_point()  
```
