---
title: "TBL Week 1"
author: "Ruoqing Zhu"
date: "Last Updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: hide
    df_print: paged
    toc: yes
    toc_float:
      collapsed: true
      smooth_scroll: true
    toc_depth: '2'
  pdf_document:
    toc: yes
    toc_depth: 2
---

```{r set-options, echo=FALSE, cache=FALSE}
  options(width = 1000)
  knitr::opts_chunk$set(fig.width=9, fig.height=7, out.width = "75%", fig.align = 'center')
  knitr::opts_chunk$set(class.source = "fold-show")
  knitr::opts_chunk$set(collapse=TRUE)
```

## Load Data

Readin the birthrate data using the code in the R-Intro lecture note. Use the data for TBL. 

```{r}
    # read-in data
    heart = read.csv("processed_cleveland.csv")
    head(heart)
```

## Question 1

Let's just observe the first 40 observations of the Age variable. Perform a hypothesis test for whether the mean age is greater than 50.

```{r}
  t.test(heart$age[1:40], mu = 50, alternative= "greater")
  t.test(heart$age[1:40], mu = 50, alternative= "less")$p.value
  t.test(heart$age[1:40], mu = 60, alternative= "greater")$p.value
  t.test(heart$age[1:10], mu = 50, alternative= "greater")$p.value
```

## Question 2

Let's use all observations of the chol variable. Perform a hypothesis test for whether the mean chol is different than 250. 

```{r}
  t.test(heart$chol, mu = 250, alternative = "two.sided")
  t.test(heart$chol, mu = 250, alternative = "greater")$p.value
  t.test(heart$chol, mu = 250, alternative = "less")$p.value
  t.test(heart$chol, mu = 50)$p.value
```


## Question 3

Test whether the mean chol is different across different gender. What is the p value? Use the equal variance assumption. 

```{r}
  male_chol = heart$chol[heart$sex == 1]
  female_chol = heart$chol[heart$sex == 0]
  t.test(male_chol, female_chol, alternative = "two.sided", var.equal = TRUE)
  
  t.test(male_chol, female_chol, alternative = "greater", var.equal = TRUE)$p.value
  t.test(male_chol, female_chol, alternative = "less", var.equal = TRUE)$p.value
```

## Question 4

First, using all observations, create a dummy variable from the heart data

  * Whether `chol` is greater than 250

Test the association between this variable and sex. What is the $p$ value?

```{r class.source = NULL, eval = TRUE}
  chol_new = (heart$chol > 250)
  chisq.test(chol_new, heart$sex)
```